---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
#### 1. Load the data
```{r, echo=TRUE}
unzip("activity.zip", overwrite = TRUE)
data <- read.csv("activity.csv", stringsAsFactors = FALSE)
```

#### 2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r, echo=TRUE}
## as.Date: converts charcters to Date using "yyyy-mm-dd" format 
data$date <- as.Date(data$date, format = "%Y-%m-%d")
```



## What is mean total number of steps taken per day?
#### 1. Make a histogram of the total number of steps taken each day
```{r, echo=TRUE}
## compute the total number of steps per day
stepsperday <- aggregate(steps ~ date, data = data, FUN = sum)
## Make a histogram
hist(stepsperday$steps, col="blue", main = "Total number of steps taken each day", xlab = "Number of steps")

```

#### 2. Calculate and report the mean and median total number of steps taken per day

the mean of total number of steps taken per day is `r mean(x = stepsperday$steps)`

the median of total number of steps taken per day is `r median(x = stepsperday$steps)`



## What is the average daily activity pattern?
#### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, echo=TRUE}
## compute the average number of steps taken across all days
asteps <- aggregate(steps ~ interval, data = data, FUN = "mean")
## create time column by adding coverting interval to H:m
asteps$time <- paste(asteps$interval%/%100, asteps$interval%%100, sep = ":")
## as.POSIXlt: converts charcters to Time using "H:M" format
asteps$time <- as.POSIXlt(asteps$time, format = "%H:%M")
plot(x = asteps$time, y = asteps$steps, type ="l", xlab = "Time", ylab = "Average number of steps", col = "blue")
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo=TRUE}
## order average steps across all the days 'asteps' by steps, extract the row of the maximum number of steps (first line)
maxstepsrow <- with(asteps, order(-steps))[1]
## extract the value of interval column for the maxstepsrow 
maxinterval <- asteps[maxstepsrow, "interval"]
```

The 5-minute interval that contain the maximum number of steps , on average across all the days in the dataset, is `r maxinterval` 



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
