---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
#### 1. Load the data
```{r, echo=TRUE}
unzip("activity.zip", overwrite = TRUE)
data <- read.csv("activity.csv", stringsAsFactors = FALSE)
```

#### 2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r, echo=TRUE}
## as.Date: converts charcters to Date using "yyyy-mm-dd" format 
data$date <- as.Date(data$date, format = "%Y-%m-%d")
```



## What is mean total number of steps taken per day?
#### 1. Make a histogram of the total number of steps taken each day
```{r, echo=TRUE}
## compute the total number of steps per day ignoring the missing values
stepsperday <- aggregate(steps ~ date, data = data, FUN = sum)
## Make a histogram
hist(stepsperday$steps, col="blue", main = "Total number of steps taken each day", xlab = "Number of steps")

```

#### 2. Calculate and report the mean and median total number of steps taken per day
```{r, echo=TRUE}
meansteps <- mean(x = stepsperday$steps)
mediansteps <- median(x = stepsperday$steps)
```


the mean of total number of steps taken per day is `r meansteps`

the median of total number of steps taken per day is `r mediansteps`



## What is the average daily activity pattern?
#### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, echo=TRUE}
## compute the average number of steps taken across all days ignoring the missing values
asteps <- aggregate(steps ~ interval, data = data, FUN = "mean")
## create time column by adding coverting interval to H:m
asteps$time <- paste(asteps$interval%/%100, asteps$interval%%100, sep = ":")
## as.POSIXlt: converts charcters to Time using "H:M" format
asteps$time <- as.POSIXlt(asteps$time, format = "%H:%M")
plot(x = asteps$time, y = asteps$steps, type ="l", xlab = "Time", ylab = "Average number of steps", col = "blue")
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo=TRUE}
## order average steps across all the days 'asteps' by steps, extract the row of the maximum number of steps (first line)
maxstepsrow <- with(asteps, order(-steps))[1]
## extract the value of interval column for the maxstepsrow 
maxinterval <- asteps[maxstepsrow, "interval"]
```

The 5-minute interval that contain the maximum number of steps , on average across all the days in the dataset, is `r maxinterval` 



## Imputing missing values
#### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r, echo=TRUE}
## the sum of na values returns the total number of missing values in the dataset for each column
## the first column steps contains all the missing values
missingvalues <- sapply(data, function(x) sum(is.na(x)))

nbna <- missingvalues[1]
```
the total number of missing values in the dataset is `r nbna`

#### 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r, echo=TRUE}
## use the mean for that 5-minute interval ignoring the missing values
avsteps <- aggregate(steps ~ interval, data = data, FUN = "mean")
## Merge the aggregate data into the original data as a new imputation column based on interval link
data1 <- merge(x = data, y = avsteps, "interval")
## rename columns
colnames(data1) <- c("interval", "steps", "date", "avsteps")

```
#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r, echo=TRUE}
## replace missing values by average value
data1$steps[is.na(data1$steps)] <- data1$avsteps[is.na(data1$steps)]
## keep usefull columns
data1 <- data1[, c("interval", "steps", "date")]
```

#### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
```{r, echo= TRUE}
## compute the total number of steps per day ignoring the missing values
stepsperday1 <- aggregate(steps ~ date, data = data1, FUN = sum)
## Make a histogram
hist(stepsperday1$steps, col="blue", main = "Total number of steps taken each day", xlab = "Number of steps")

meansteps1 <- mean(x = stepsperday1$steps)
mediansteps1 <- median(x = stepsperday1$steps)
```
the mean of total number of steps taken per day is `r meansteps1`

the median of total number of steps taken per day is `r mediansteps1`

the difference of mean of total number of steps taken per day is `r meansteps1 - meansteps`

the difference of median of total number of steps taken per day is `r mediansteps1 - mediansteps`

the impact of imputing missing data on the estimates of the total daily number of steps is median and mean are the same.

## Are there differences in activity patterns between weekdays and weekends?
